---
title: "Final Project"
output: html_document
---

First will look at salaries over time.
```{r,echo=FALSE}
library(tidyverse)
setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
ppos <- read.csv("Appearances.csv",stringsAsFactors=FALSE)
batting <- read.csv("Batting.csv",stringsAsFactors=FALSE)
salaries <- read.csv("Salaries.csv",stringsAsFactors=FALSE)
```
Note that salary information is only available from 1985 forward.
```{r}
min(salaries$yearID)
```

We can observe salary trends over time using player salary data from the Lahman dataset, and US household salary data from the US Census Bureau.
```{r}

```
Using the respective 1985 salaries as a baseline, we see that median player salary growth has not outpaced the salary growth of US households. W

Since 1985, US median household income has increased by around 25% every 5 years.
Baseball salary fluctuated through the late 1980s and 90s, but increased dramatically after 2000. Was this related to a performance boost? Average salary has been increasing at a much faster pace. This suggests that the salary for top players is increasing faster than that of "average" players.