---
title: "Final Project"
output: html_document
---

First will look at salaries over time.
```{r,warning=FALSE,echo=FALSE,message=FALSE}
library(tidyverse)
library(gridExtra)
setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
salaries <- read.csv("Salaries.csv",stringsAsFactors=FALSE)
```
Note that salary information is only available from 1985 forward.
```{r}
min(salaries$yearID)
```

We can observe salary trends over time using player salary data from the Lahman dataset, and US household salary data from the US Census Bureau.

```{r,warning=FALSE,echo=FALSE}
BBsalaries = salaries %>% group_by(yearID) %>% summarize(msalary=median(salary),avg=mean(salary), q25 = quantile(salary,0.25), q75 = quantile(salary,0.75), q90 = quantile(salary,0.9))
setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
USsalaries <- read.csv("USmedsalary.csv")
USsalaries <- USsalaries %>% filter(yearID > 1984)
USsalaries <- USsalaries[order(USsalaries$yearID),]

BBsalaries$BBpct <- BBsalaries$msalary/BBsalaries$msalary[[1]]
BBsalaries$BBavgpct <- BBsalaries$avg/BBsalaries$avg[[1]]
USsalaries$USpct <- USsalaries$msalary/USsalaries$msalary[[1]]
USsalaries$USavgpct <- USsalaries$avg/USsalaries$avg[[1]]

allsal <- merge(BBsalaries,USsalaries,by="yearID")

override.color <- c("red","red","blue","blue")
override.linetype <- c(1,2,1,2)

ggplot(allsal,aes(x=yearID)) + 
  geom_line(aes(y=USpct,color="US median",linetype="US median")) +
  geom_line(aes(y=USavgpct,color="US average",linetype="US average")) +
  geom_line(aes(y=BBpct,color="MLB median",linetype="MLB median")) +
  geom_line(aes(y=BBavgpct,color="MLB average",linetype="MLB average")) +
  scale_x_continuous(breaks = seq(1985,2015,5)) +
  scale_y_continuous(breaks = seq(0,10,1)) +
  labs(x="Year",y="Salary level vs 1985 (multiplier)",title="MLB salary vs US household income") +
  scale_color_manual(values = c("red","red","blue","blue")) +
  scale_linetype_manual(values = c(1,2,1,2)) +
  guides(color = guide_legend(override.aes = list(color = override.color, linetype = override.linetype),title="")) +
  guides(linetype=FALSE)

```

Using the respective 1985 salaries as a baseline, we see that US median household income has increased by around 25% every 5 years. Median player salary growth generally not outpaced the salary growth of US households until the past decade. Average salary, in contrast, has steadily increased since 1990. This suggests that salary trends for "top" players may be different from that of "average" players.

```{r,echo=FALSE}
override.color <- c("red","red","seagreen3","blue")
override.linetype <- c(2,1,2,1)

ggplot(allsal,aes(x=yearID)) + 
  geom_line(aes(y=q25/1000000,color="25th",linetype="25th")) +
  geom_line(aes(y=msalary.x/1000000,color="50th",linetype="50th")) +
  geom_line(aes(y=q75/1000000,color="75th",linetype="75th")) +
  geom_line(aes(y=q90/1000000,color="90th",linetype="90th")) +
  scale_x_continuous(breaks = seq(1985,2015,5)) +
  scale_y_continuous(breaks = seq(0,15,2.5)) +
  labs(x="Year",y="Salary ($, millions)",title="MLB salary growth, percentiles") +
  scale_color_manual(values = c("red","red","seagreen3","blue")) +
  scale_linetype_manual(values = c(2,1,2,1)) +
  guides(color = guide_legend(override.aes = list(color = override.color, linetype = override.linetype),title="")) +
  guides(linetype=FALSE)
```

Observing the trends by percentile, we can see that, indeed, salaries have not increased across the board, with large gains for the top paid players, with little change for players in the lower percentiles. We will look at a comparison of percentiles by team.

```{r,echo=FALSE,warning=FALSE}
salteam <- salaries %>% select(playerID,yearID,teamID,salary) %>% na.omit() %>% group_by(yearID,teamID) %>% summarize(totsal = sum(salary), p50sal = quantile(salary,.5), p90sal = quantile(salary,.9), p90rat = quantile(salary,.9)/median(salary), p10sal = quantile(salary,.9)/quantile(salary,.1))

override.color = c("seagreen3","orangered1")
salteam %>% ggplot() + 
  geom_boxplot(aes(yearID,p50sal/1000000,factor=factor(yearID),color="50th"),outlier.shape=NA, na.rm = TRUE) +
  geom_boxplot(aes(yearID,p90sal/1000000,factor=factor(yearID),color="90th"),outlier.shape=NA, na.rm = TRUE) +
  scale_x_continuous(breaks = seq(1985,2015,5)) +
  scale_y_continuous(trans="log10") +
  scale_color_manual(values = c("seagreen3","orangered1")) +
  labs(x = "Year", y = "Salary ($, millions)", title = "Salaries across teams") +
  guides(color = guide_legend(override.aes = list(color = override.color),title="Percentile"))
```

On a team-by-team basis, we can see that salary growth was not the same between the median and the 90th salary percentiles. In particular, there was a decreasing trend in median salary between 1991 and 1995, where the 90th percentile did not change. What may have caused this?

```{r,echo=FALSE,warning=FALSE}
salteam %>% ggplot(aes(yearID,p90rat,fill=factor(yearID))) + 
  geom_boxplot(na.rm = TRUE) + theme(legend.position="none") + 
  scale_x_continuous(breaks = seq(1985,2015,5)) +
  scale_y_continuous(breaks = seq(0,30,5)) +
  geom_vline(aes(xintercept=as.numeric(1994)),linetype=1,size=5,alpha=0.5, color="red") +
  geom_text(aes(x=1994,y=25),label="STRIKE",angle=90,hjust = 0,color="white") +
  labs(x="Year",y="90th percentile รท median salary")
```

Using the same data to plot the [greater percentage the 90th percentile was receiving], we can see an increasing trend until 1995, the year after the baseball strike of 1994, then a decreasing trend afterwards. Again, percentiles were calculated by team. [Validate with RDD?]

[rdd]

In 1994, two-pronged plan to limit the effect of reduced revenues. First was to enact revenue-sharing in order to keep smaller teams from having to drop out. The other was a salary cap, to limit the labor cost of players. Though the salary thing maybe didn't go through, revenue sharing did. Was it effective?

```{r,echo=FALSE,warning=FALSE}
setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
revs <- read.csv("MLBrevs.csv",stringsAsFactors=FALSE)
names <- read.csv("Teamnames.csv",stringsAsFactors=FALSE)
cpi <- read.csv("cpi.csv")

# Generate franchise ID for each team in "revs" data gathered from Forbes; accounts for team name changes and standardizes to Lahman franchise notation
revs$franchID = ""
for (i in 1:nrow(revs)){
  nameyear <- names %>% filter(yearID==revs$Year[[i]])
  nameab <- nameyear %>% filter(grepl(revs$Team[[i]],nameyear$name)==TRUE)
  revs$franchID[[i]] = nameab$franchID[[1]]
}

# Calculate inflation factor by year vs 1990 dollars (v1990)
cpi <- cpi %>% filter(year>1989)
cpi1990 = cpi$cpi[[1]]
cpi <- cpi %>% mutate(v1900 = cpi/cpi1990)
for (i in 1:nrow(revs)){
  revs$v1990[[i]] = cpi$v1900[which(cpi$year==revs$Year[[i]])]  
}

ggplot(revs) + geom_boxplot(aes(Year,Revenue/v1990,fill=factor(Year))) +
  geom_vline(aes(xintercept=as.numeric(1994)),linetype=1,size=5,alpha=0.5, color="red") +
  geom_text(aes(x=1994,y=200),label="STRIKE",angle=90,hjust = 0,color="white") +
  labs(x="Year",y="Revenue in 1990 dollars, millions") +
  scale_x_continuous(breaks = seq(1990,2015,5)) +
  scale_y_continuous(breaks = seq(0,300,50)) +
  theme(legend.position="none")
```

We see that there is a drop in revenues in the year of the stike, but this recovers after a few years. Note that 2012 revenue data is incomplete and skewed upwards.

```{r,echo=FALSE}
ggplot(revs) + geom_boxplot(aes(Year,Value/v1990,fill=factor(Year))) +
  geom_vline(aes(xintercept=as.numeric(1994)),linetype=1,size=5,alpha=0.5, color="red") +
  geom_text(aes(x=1994,y=500),label="STRIKE",angle=90,hjust = 0,color="white") +
  labs(x="Year",y="Valuation in 1990 dollars, log millions") +
  scale_x_continuous(breaks = seq(1990,2015,5)) +
  scale_y_continuous(breaks = c(125,250,500,1000,2000),trans="log10") +
  theme(legend.position="none")
```

Team valuations rose greatly after the revenue sharing agreement, and again after 2010. Let's look back at revenues. 2012 valuation data is complete.

```{r,echo=FALSE}
revs <- revs[order(revs$franchID,revs$Year),]
revs$Value = as.numeric(revs$Value)

for (i in 2:nrow(revs)){
  if (revs$Year[[i]]>1990){
    revs$dvpct[[i]] = (revs$Value[[i]]-revs$Value[[i-1]])/revs$Value[[i-1]]
    revs$dv[[i]] = (revs$Value[[i]]-revs$Value[[i-1]])
    revs$dr[[i]] = (revs$Revenue[[i]]-revs$Revenue[[i-1]])/revs$Revenue[[i-1]]
    revs$di[[i]] = (revs$O_Income[[i]]-revs$O_Income[[i-1]])/revs$O_Income[[i-1]]
  }
}

revs %>% filter(Year > 1990) %>% ggplot() + geom_boxplot(aes(Year,dvpct,fill=factor(Year))) +
  geom_vline(aes(xintercept=as.numeric(1994)),linetype=1,size=7,alpha=0.5, color="red") +
  geom_text(aes(x=1994,y=0.5),label="STRIKE",angle=90,hjust = 0,color="white") +
  labs(x="Year",y="Valuation in 1990 dollars, log millions") +
  scale_x_continuous(breaks = seq(1990,2015,5)) +
  #scale_y_continuous(breaks = c(125,250,500,1000,2000),trans="log10") +
  theme(legend.position="none")
```



```{r,echo=FALSE}
revs$deval <- ifelse(revs$dv<0,1,0)

base.plot <- ggplot() +
  scale_x_continuous(limits = c(1989,2016), breaks=seq(1990,2015,5)) +
  scale_fill_distiller(palette = "Spectral",breaks=seq(0,70,20),name="Percent") +
  geom_vline(aes(xintercept=1993.5),linetype=3, color="black") +
  geom_vline(aes(xintercept=1994.5),linetype=3, color="black") +
  geom_vline(aes(xintercept=2001.5),linetype=3, color="black") +
  geom_vline(aes(xintercept=2006.5),linetype=3, color="black") +
  geom_vline(aes(xintercept=2010.5),linetype=3, color="black")

revs$oloss <- ifelse(revs$O_Income<0,1,0)
oloss <- revs %>% group_by(Year) %>% summarize(teams = sum(oloss)/n())
oloss.plot <- base.plot + geom_bar(aes(Year,teams*100,fill=teams*100),stat="identity",data = oloss) + 
  theme(axis.title.x = element_blank(), axis.text.x = element_blank()) +
  scale_y_continuous(breaks=seq(0,80,10),limits=c(0,80)) + 
  geom_text(aes(x=1994,y=5),label="STRIKE",angle=90,hjust = 0,color="white") +
  geom_text(aes(x=1990.5,y=80),label="CBA7",hjust = 0,color="black",size=3.5) +
  geom_text(aes(x=1997,y=80),label="CBA8",hjust = 0,color="black",size=3.5) +
  geom_text(aes(x=2003,y=80),label="CBA9",hjust = 0,color="black",size=3.5) +
  geom_text(aes(x=2007.2,y=80),label="CBA10",hjust = 0,color="black",size=3.5) +
  geom_text(aes(x=2012,y=80),label="CBA11",hjust = 0,color="black",size=3.5) +
  labs(y="Teams with operating loss (%)")

deval <- revs %>% filter(Year>1990) %>% group_by(Year) %>% summarize(teams = sum(deval)/n())
deval.plot <- base.plot + geom_bar(aes(Year,teams*100,fill=teams*100),stat="identity",data=deval) +
  scale_y_continuous(breaks=seq(0,80,10)) +
  labs(y="Teams that lost value (%)")

grid.arrange(oloss.plot,deval.plot,ncol=1)
```

regress/anova by cba?
```{r}
summary(lm(teams ~ Year,data=oloss))
summary(lm(teams ~ Year,data=deval))
```

```{r}
addcba <- function(df){
  df$cba = .bincode(df$Year,c(0,1993.5,1994.5,2002,2006,2010,10000))
  df$cba = ifelse(df$cba>2,df$cba+5,df$cba)
  df$cba = ifelse(df$cba==1,df$cba+6,df$cba)
  df$cba = ifelse(df$cba==2,NA,df$cba)
  df
}
oloss = addcba(oloss)
deval = addcba(deval)

summary(lm(teams ~ cba,data=oloss))
summary(lm(teams ~ cba,data=deval))
```


Measures seem to prevent teams from losing value year-on-year.

https://www.sbnation.com/2010/8/30/1065675/8-30-2002-baseball-avoids-another
2002 - new collective bargaining agreement that increasing rev sharing 70% and added luxury tax

https://www.fangraphs.com/tht/a-history-of-the-collective-bargaining-agreement-part-3/
^good

So it seems like the measures accomplish the goal of keeping teams on good financial footing. Let's see how they affect team composition.

WAR is a stat and blah blah blah
```{r,echo=FALSE,eval=FALSE}
# Add salary and other numbers to WAR table
1 = 1 #prevent running accidentally since this takes a while to run

library(tidyverse)
setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017/war_archive-2017-11-08")
wars <- read.csv("war_daily_bat.csv",stringsAsFactors=FALSE)
setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
salaries <- read.csv("Salaries.csv",stringsAsFactors=FALSE)
#bats <- read.csv("batting.csv")

salaries <- salaries %>% filter(yearID > 1969 & yearID < 2017)
wars <- wars %>% filter(year_ID > 1969)
wars <- wars %>% rename(playerID=player_ID,yearID=year_ID,teamID=team_ID)

salaries <- salaries[order(salaries$playerID,salaries$yearID),]
wars <- wars[order(wars$playerID,wars$yearID),]

#Determine salary bumps/reductions in WAR dataset
players <- unique(wars$playerID)
wars$salary[which(wars$salary=="NULL")] <- "-999"
wars$salary <- as.numeric(wars$salary)

wars$bump = 0
wars$cut = 0
wars$dsal = 0

wars$WAA = as.numeric(wars$WAA)
wars$WAA_off = as.numeric(wars$WAA_off)
wars$WAA_def = as.numeric(wars$WAA_def)
wars$WAR = as.numeric(wars$WAR)
wars$WAR_off = as.numeric(wars$WAR_off)
wars$WAR_def = as.numeric(wars$WAR_def)

player1 <- wars$playerID[[1]]
for (i in 2:nrow(wars)) {
  if (wars$playerID[[i]] == player1){
    wars$prevWAA[[i]] = wars$WAA[[i-1]]
    wars$dWAA[[i]] = wars$WAA[[i]] - wars$WAA[[i-1]]
    wars$prevWAA_off[[i]] = wars$WAA_off[[i-1]]
    wars$dWAA_off[[i]] = wars$WAA_off[[i]] - wars$WAA_off[[i-1]]
    wars$prevWAA_def[[i]] = wars$WAA_def[[i-1]]
    wars$dWAA_def[[i]] = wars$WAA_def[[i]] - wars$WAA_def[[i-1]]
    
    wars$prevWAR[[i]] = wars$WAR[[i-1]]
    wars$dWAR[[i]] = wars$WAR[[i]] - wars$WAR[[i-1]]
    wars$prevWAR_off[[i]] = wars$WAR_off[[i-1]]
    wars$dWAR_off[[i]] = wars$WAR_off[[i]] - wars$WAR_off[[i-1]]
    wars$prevWAR_def[[i]] = wars$WAR_def[[i-1]]
    wars$dWAR_def[[i]] = wars$WAR_def[[i]] - wars$WAR_def[[i-1]]
    
    if (wars$salary[[i]]==-999){
      wars$salary[[i]]=wars$salary[[i-1]]
    }
    #bump if salary increased
    if (wars$salary[[i]] > wars$salary[[i-1]]){
      wars$bump[[i]] = 1
    }
    #cut if salary decreased
    else if (wars$salary[[i]] < wars$salary[[i-1]]){
      wars$cut[[i]] = 1
    }
    wars$dsal[[i]] = wars$salary[[i]] - wars$salary[[i-1]]
    wars$pdsal[[i]] = (wars$salary[[i]] - wars$salary[[i-1]])/wars$salary[[i-1]]
  }
  else{
    player1 <- wars$playerID[[i]]
    if (wars$salary[[i]]==-999){
      wars$salary[[i]]=0
    }
  }
}

setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
write.csv(wars,"warplus.csv")
```

```{r,echo=FALSE,warning=FALSE}
library(tidyverse)
setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
wars <- read.csv("warplus.csv",stringsAsFactors=FALSE)
names <- read.csv("Teams.csv",stringsAsFactors=FALSE)

names <- names %>% select(yearID,teamID,teamIDBR,teamIDlahman45,teamIDretro,franchID,W,L,Rank)
names <- names %>% filter(yearID>1984 & yearID<2017)
```

```{r,echo=FALSE,warning=FALSE}
wars <- wars %>% filter(yearID>1984 & yearID<2017)
wars$OPS_plus <- as.numeric(wars$OPS_plus)
wars$WAR <- as.numeric(wars$WAR)
wars$WAR_off <- as.numeric(wars$WAR_off)

warteam <- wars %>% na.omit() %>% group_by(yearID,teamID) %>% 
  summarize(medOPS = median(OPS_plus),
            q90OPS = quantile(OPS_plus,0.9),
            q75OPS = quantile(OPS_plus,0.75),
            q25OPS = quantile(OPS_plus,0.25),
            q10OPS = quantile(OPS_plus,0.1),
            sdOPS = sd(OPS_plus),
            medWAR = median(WAR),
            q90WAR = quantile(WAR,0.9),
            q75WAR = quantile(WAR,0.75),
            q25WAR = quantile(WAR,0.25),
            q10WAR = quantile(WAR,0.1),
            sdWAR = sd(WAR),
            medWAR_off = median(WAR_off),
            q90WAR_off = quantile(WAR_off,0.9),
            q75WAR_off = quantile(WAR_off,0.75),
            q10WAR_off = quantile(WAR_off,0.1),
            WARtot = sum(WAR),
            salteam = sum(salary))

warteam %>% ggplot(aes(yearID,medWAR,fill=factor(yearID))) + 
  geom_boxplot(na.rm = TRUE) + theme(legend.position="none") + 
  scale_x_continuous(breaks = seq(1985,2015,5)) + 
  labs(y = "Median WAR (by team)")

```


```{r,echo=FALSE}
warteam$franchID = ""
warteam$wlratio = 0
warteam$rank = 0
for (i in 1:nrow(warteam)){
  nameyear <- names %>% filter(yearID==warteam$yearID[[i]])
  nameab <- nameyear %>% filter(nameyear$teamIDBR==warteam$teamID[[i]])
  warteam$franchID[[i]] = nameab$franchID[[1]]
  warteam$wlratio[[i]] = nameab$W[[1]]/nameab$L[[1]]
  warteam$rank[[i]] = nameab$Rank[[1]]
}

#order franchises by total WAR prior to the year 2000
allwar <- warteam %>% filter(yearID<2000)
allwar <- allwar %>% group_by(franchID) %>% summarize(wartot=sum(medWAR))
allwar <- allwar %>% arrange(wartot) #ascending
#allwar <- allwar %>% arrange(desc(wartot)) #descending
allwar$rank <- c(1:nrow(allwar))
allwar$franfct <- allwar$rank
allwar$franfct <- factor(allwar$franfct,levels=allwar$rank,labels=allwar$franchID)

warteam$franfct <- allwar$franfct[[1]]
for (i in 1:nrow(warteam)){
  warteam$franfct[[i]] <- allwar$franfct[which(allwar$franchID==warteam$franchID[[i]])]
}
```

add vline, change palette
```{r,echo=FALSE}
jet.colors <- colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan", "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))

warteam %>% ggplot() + geom_tile(aes(x=yearID,y=factor(franfct),fill=medWAR)) + 
  scale_x_continuous(breaks = seq(1985,2015,5)) + 
  #scale_y_discrete(labels = rev(allwar$franchID)) +
  scale_fill_gradientn(colors = jet.colors(10),name="Median WAR\n   (offense)") +
  labs(x="Year",y="Team/Franchise")

warteam %>% ggplot() + geom_tile(aes(x=yearID,y=factor(franfct),fill=q75WAR)) + 
  scale_x_continuous(breaks = seq(1985,2015,5)) + 
  #scale_y_discrete(labels = rev(allwar$franchID)) +
  scale_fill_gradientn(colors = jet.colors(10),name="75th pctile WAR\n     (offense)") +
  labs(x="Year",y="Team/Franchise")
```

The teams with the highest WAR prior to 2000 

```{r,echo=FALSE}
warteam %>% ggplot() + geom_tile(aes(x=yearID,y=factor(franfct),fill=wlratio)) + 
  scale_x_continuous(breaks = seq(1985,2015,5)) + 
  #scale_y_discrete(labels = rev(allwar$franchID)) +
  scale_fill_gradientn(colors = jet.colors(10), name="wlS") +
  labs(x="Year",y="Team/Franchise")

warteam %>% ggplot() + geom_tile(aes(x=yearID,y=factor(franfct),fill=rank)) + 
  scale_x_continuous(breaks = seq(1985,2015,5)) + 
  #scale_y_discrete(labels = rev(allwar$franchID)) +
  scale_fill_gradientn(colors = jet.colors(10), name="wlS") +
  labs(x="Year",y="Team/Franchise")
```

WL ratio and rank were not affected, random as always.

What kind of players were paid more?

[1995]
```{r,warning=FALSE,echo=FALSE,message=FALSE}
salmax95 <- salaries %>% filter(yearID==1995) %>% select(playerID,teamID,salary) %>% na.omit() %>% group_by(teamID) %>% filter(salary == max(salary))
salmax95 <- unique(salmax95)

setwd("/Users/Chris/Dropbox/School Stuff/BST 260 Data Science/baseball2017")
players <- read.csv("Master.csv",stringsAsFactors=FALSE)
playernames <- players %>% filter(playerID %in% salmax95$playerID) %>% select(playerID,nameFirst,nameLast)
salmax95 <- merge(salmax95,playernames,by="playerID")
salmax95$fullname <- with(salmax95, paste0(nameFirst," ",nameLast))
salmax95 <- salmax95[order(salmax95$salary),]

library(ggrepel)
salmax95 %>% ggplot(aes(x=1995,y=salary/1000000,label=nameFirst)) +
  geom_text_repel(aes(label=fullname),segment.color = 'grey3',segment.alpha=0.25,label.padding=1.5) + 
  geom_point(aes(color=salary/1000000)) +
  scale_color_gradientn(colors = rainbow(5),breaks=seq(0,10,2),name="Percent") +
  scale_x_continuous(breaks=c()) +
  labs(title="Highest-paid players per team, 1995",x="",y = "Salary ($, millions)")
```

Max salaries in a team went to future Hall of Famers Ken Griffey Jr., Barry Larkin, Kirby Puckett, and Cal Ripken, Jr., as well as other popular players of the time, including Barry Bonds, Jose Canseco, Mark McGwire, and Daryl Strawberry. So, it seems that salaries went to popular players, who were presumably also the players who played well.

Do better players make more?

```{r}
#Previously calculated salary changes and performance changes year-to-year and saved in warplus.csv
warstats <- wars %>% filter(pitcher=="N")
warstats <- warstats[c("playerID","yearID","teamID",
                       "WAA","WAA_off","WAA_def",
                       "WAR","WAR_off","WAR_def",
                       "prevWAA","prevWAA_off","prevWAA_def",
                       "prevWAR","prevWAR_off","prevWAR_def",
                       "dWAA","dWAA_off","dWAA_def",
                       "dWAR","dWAR_off","dWAR_def",
                       "OPS_plus","salary","bump","cut","dsal","pdsal")]
warstats <- warstats %>% filter(yearID > 1985)

warstats <- warstats %>% filter(dsal!=0)
warstats <- warstats %>% filter(abs(dsal)>1)

warzeros <- warstats %>% filter(dsal==0)
warzeros %>% ggplot() + geom_histogram(aes(as.numeric(prevWAA_off)),binwidth=0.05)

warcut <- warstats %>% filter(dsal<0)
warcut %>% ggplot() + geom_histogram(aes(as.numeric(prevWAA_off)),binwidth=0.05)

warbump <- warstats %>% filter(dsal>0)
warbump %>% ggplot() + geom_histogram(aes(as.numeric(prevWAA_off)),binwidth=0.05)
```

Doesn't say much. Let's look at other trends.

```{r}
warstats$pdsal[which(warstats$pdsal==Inf)] = NA

warstats %>% filter(pdsal < 40, dsal > -100000) %>% ggplot() + 
  geom_point(aes(prevWAA_off,dsal + abs(min(dsal))+1 )) +
  geom_smooth(aes(prevWAA_off,dsal + abs(min(dsal))+1 )) +
  scale_y_continuous(trans="log10")

warstats %>% ggplot() + 
  geom_point(aes(prevWAA_off,dsal/1000000)) +
  geom_smooth(aes(prevWAA_off,dsal/1000000)) +
  scale_y_continuous(breaks=seq(-30,30,5)) +
  labs(x="WAA in previous year", y="Salary change ($, millions)")

warstats %>% ggplot() + 
  geom_point(aes(prevWAR_off,dsal/1000000)) +
  geom_smooth(aes(prevWAR_off,dsal/1000000)) +
  scale_y_continuous(breaks=seq(-30,30,5)) +
  labs(x="WAR in previous year", y="Salary change ($, millions)")

warstats %>% ggplot() + 
  geom_point(aes(prevWAR,dsal/1000000)) +
  geom_smooth(aes(prevWAR,dsal/1000000)) +
  scale_y_continuous(breaks=seq(-30,30,5)) +
  labs(x="WAR in previous year", y="Salary change ($, millions)")

# warbump %>% filter(pdsal < 10000) %>% ggplot() + 
#   scale_y_continuous(trans="log10") +
#   geom_point(aes(as.numeric(prevWAA_off),pdsal))
# warcut %>% filter(pdsal < 10000) %>% ggplot() + 
#   geom_point(aes(as.numeric(prevWAA),pdsal))
```

Again, doesn't look like much. Let's look at the regression.

```{r}
w2 <- warstats %>% filter(yearID==1991)
#get est and r^2 by year?
summary(lm(w2$dsal~w2$prevWAA_off))
summary(lm(w2$dsal~w2$prevWAR_off))
summary(lm(w2$dsal~w2$prevWAR))
summary(lm(w2$dsal ~ w2$prevWAR + w2$dWAR) )
```

Sure enough, doesn't look like much. But what if it changes over time?

```{r}
w2 <- warstats

library(broom)
regres <- w2 %>% group_by(yearID) %>% do(glance(lm(dsal ~ prevWAR,data=.)))

regres %>% ggplot() + geom_bar(aes(yearID,adj.r.squared),stat="identity")
```

```{r}
w2 <- warstats

library(broom)
regres <- w2 %>% group_by(yearID) %>% do(glance(lm(salary ~ WAR,data=.)))

regres %>% ggplot() + geom_bar(aes(yearID,adj.r.squared,fill=adj.r.squared),stat="identity") +
    scale_x_continuous(limits = c(1984,2016), breaks=seq(1985,2015,5)) +
    scale_fill_distiller(palette = "Spectral",breaks=seq(0,70,20),name="Percent") +
    geom_text(aes(x=1994,y=0.02),label="STRIKE",angle=90,hjust = 0,color="white") +
    geom_text(aes(x=1986,y=.4),label="CBA6",hjust = 0,color="black",size=3.5) +
    geom_text(aes(x=1990.5,y=.4),label="CBA7",hjust = 0,color="black",size=3.5) +
    geom_text(aes(x=1997,y=.4),label="CBA8",hjust = 0,color="black",size=3.5) +
    geom_text(aes(x=2003,y=.4),label="CBA9",hjust = 0,color="black",size=3.5) +
    geom_text(aes(x=2007.2,y=.4),label="CBA10",hjust = 0,color="black",size=3.5) +
    geom_text(aes(x=2012,y=.4),label="CBA11",hjust = 0,color="black",size=3.5) +
    geom_vline(aes(xintercept=1989.5),linetype=3, color="black") +
    geom_vline(aes(xintercept=1993.5),linetype=3, color="black") +
    geom_vline(aes(xintercept=1994.5),linetype=3, color="black") +
    geom_vline(aes(xintercept=2001.5),linetype=3, color="black") +
    geom_vline(aes(xintercept=2006.5),linetype=3, color="black") +
    geom_vline(aes(xintercept=2010.5),linetype=3, color="black")
```

Looks like this may have changed along with the collective bargaining agreements. May have had to do with luxury tax. Or maybe steroid era.

Do better players get awards?

```{r}

```


Predict awards


Conclusion





























