---
title: "SVM iris"
output: html_document
---
http://rischanlab.github.io/SVM.html

Use library e1071, you can install it using install.packages(“e1071”). Load library
```{r}
library(e1071)
head(iris,5)
```

Using Iris data
```{r}
attach(iris)
```

Divide Iris data to x (containt the all features) and y only the classes
```{r}
x <- subset(iris, select=-Species)
y <- Species
```

Create SVM Model and show summary
```{r}
svm_model <- svm(Species ~ ., data=iris)
summary(svm_model)
```

Run Prediction and you can measuring the execution time in R
```{r}
pred <- predict(svm_model,x)
system.time(pred <- predict(svm_model,x))
```

See the confusion matrix result of prediction, using command table to compare the result of SVM prediction and the class data in y variable.
```{r}
table(pred,y)
```

TUning SVM to find the best cost and gamma ..
```{r}
svm_tune <- tune(svm, train.x=x, train.y=y, 
              kernel="radial", ranges=list(cost=10^(-1:2), gamma=c(.5,1,2)))

print(svm_tune)
```



